---
title: "Ranked Sampling Workflow for UDKIK"
author: "Anders Bjørn Møller, SEGES"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: true
    fig_width: 10
    fig_height: 7
    keep_md: true
params:
  data_dir: "C:/Users/anbm/OneDrive - SEGES Innovation PS/UDKIK/Data/"
  study_area_idx: 3
---

## Purpose

This document presents a workflow to generate ranked sampling points. The goal is to place one primary sampling point per cluster, add alternative sampling points within the clusters, depending on their sizes, and rank the points by distance to the cluster center.

During field sampling, one sample should be collected per cluster. The primary sampling point is the first choice, and if it is not accessible, or outside the area of interest, the alternative sampling points in the same cluster should be used. The alternative sampling points should be prioritized according to their rank, indicated by a letter suffix (e.g., 7a, 7b, 7c).

## Workflow Overview

The workflow follows these stages:

1.  Read study area and raster inputs.
2.  Crop and mask rasters to one field.
3.  Transform input layers to percentiles.
4.  Clamp outliers to reduce extreme values.
5.  Build a candidate mask for valid sample locations.
6.  Run primary k-means sampling with spatial weighting.
7.  Add extra points inside large clusters.
8.  Rank points within each cluster by center distance.
9.  Plot final labeled points.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(terra)
library(samplekmeans)
library(tidyverse)
library(dplyr)
library(tidyterra)
library(vctrs)
library(rcartocolor)

set.seed(124)

index_to_letters <- function(n) {
  sapply(n, function(k) {
    out <- ""
    while (k > 0) {
      k <- k - 1
      out <- paste0(letters[(k %% 26) + 1], out)
      k <- k %/% 26
    }
    out
  })
}
```

## Parameters and Input Validation

The workflow is demonstrated for study area 3 in the UDKIK project, but it is applicable to any area within the coverage of the input rasters. In this example, the input rasters include elevation, peat probability, and the vertical distance to the channel network (vdtochn). The workflow can be adapted to other raster inputs as needed.

The parameters `sampling_zones_ha` and `sampling_points_ha` control the number of clusters and sampling locations per ha, respectively. The number of clusters is determined by the size of the study area (one cluster per ha). The number of sampling locations depends on the areas of the clusters, set to 4 sampling locations per ha in this example.

```{r preflight}
dir_data <- params$data_dir
study_area_idx <- params$study_area_idx

study_area_path <- file.path(dir_data, "Study_areas", "Study_areas.shp")
sampling_paths <- file.path(
  dir_data,
  c(
    "Sampling_input/dhm2015_terraen_10m.tif",
    "Sampling_input/peat_probability_2025_resample.tif",
    "Sampling_input/vdtochn.tif"
  )
)

sampling_input <- rast(sampling_paths)

names(sampling_input) <- basename(sampling_paths) |>
  tools::file_path_sans_ext()

sampling_zones_ha <- 1
sampling_points_ha <- 4
```

```{r stop-if-missing, include=FALSE}
checks <- tibble::tibble(
  item = c("Study areas", "Terrain", "Peat probability", "Hydrology"),
  path = c(study_area_path, sampling_paths),
  exists = file.exists(c(study_area_path, sampling_paths))
)

checks

can_run <- all(checks$exists)

if (!can_run) {
  cat("Missing one or more required data files. Update params$data_dir and knit again.\n")
}

if (!can_run) {
  knitr::knit_exit()
}
```

## 1. Study Area Selection

The study area covers 8.9 ha, and is divided into two agricultural fields. The variable `study_area_idx` is an integer for indexing the study area polygons.

```{r study-area, out.width = "60%"}
study_areas <- vect(study_area_path)

values(study_areas)

plot(study_areas[study_area_idx, ], main = paste("Study area", study_area_idx))
```

## 2. Load and Subset Raster Inputs

The input rasters are cropped and masked to the selected study area. Raster cells are omitted if they contain NA values in any of the input layers.

```{r load-rasters, out.width = "80%"}
sampling_input_field <- crop(
  sampling_input,
  study_areas[study_area_idx, ]
) |>
  mask(
    study_areas[study_area_idx, ],
    touches = FALSE
  )

sampling_input_field <- ifel(
  sum(is.na(sampling_input_field)) == 0,
  sampling_input_field,
  NA
)

names(sampling_input_field) <- names(sampling_input)

plot(sampling_input_field,
     extent = study_areas[study_area_idx, ],
  fun = function() {
    plot(study_areas[study_area_idx, ], add = TRUE)
  },
  buffer = TRUE
)
```

## 3. Percentile Transformation and Outlier Clamp

The input layers are transformed to percentiles using the empirical cumulative distribution function (ECDF). The resulting values are also clamped to three standard deviations from the mean to reduce the influence of extreme values. This is mainly relevant if any of the input layers contain many duplicate values, which can lead to large jumps in the ECDF.

The transformation into percentiles helps to make cluster sizes more uniform and counteracts the potential effects of extreme values.

```{r pctile-and-clamp, out.width = "80%"}
sampling_input_pctile <- sapp(
  sampling_input_field,
  function(x) {
    x_ecdf <- ecdf(values(x))
    app(x, x_ecdf)
  }
)

field_means <- global(sampling_input_pctile, "mean", na.rm = TRUE) |>
  unlist()
field_sds <- global(sampling_input_pctile, "sd", na.rm = TRUE) |>
  unlist()

sampling_input_pctile <- sampling_input_pctile |>
  clamp(
    lower = field_means - field_sds * 3,
    upper = field_means + field_sds * 3
  )

names(sampling_input_pctile) <- names(sampling_input_field)

plot(
  sampling_input_pctile,
  extent = study_areas[study_area_idx, ],
  fun = function() {
    plot(study_areas[study_area_idx, ], add = TRUE)
  },
  buffer = TRUE
)
```

## 4. Candidate Sampling Pixels

The selection of sampling points is restricted to areas more than 10 m from the field boundaries. All pixels in the input are used for identifying clusters, but sampling locations are limited to the candidate pixels.

```{r candidates, out.width = "60%"}
candidates_field <- ifel(
  is.na(sum(sampling_input_field)),
  NA,
  1
) |>
  mask(
    study_areas[study_area_idx, ] |>
      buffer(width = -10),
    touches = FALSE
)

plot(
  candidates_field, main = "Candidate sampling pixels",
  extent = study_areas[study_area_idx, ],
  buffer = TRUE
)
plot(study_areas[study_area_idx, ], add = TRUE)
```

## 5. Primary Cluster Sampling

The sampling zones and primary sampling locations are generated based on k-means clustering, using the package `samplekmeans`. The number of clusters is determined by the size of the study area, with one cluster per ha (rounded off).

The parameter `xy_weight` is set to 2, which means that the spatial coordinates are weighted twice as much as the input raster values. This encourages clusters to be more spatially compact, while still considering the input raster values. The `candidates` parameter ensures that only valid sampling locations are selected, and the `min_cluster_size` parameter ensures that clusters with fewer than 50 pixels (i.e. areas less than 0.5 ha) are not produced.

```{r primary-clustering, out.width = "60%"}
myclusters <- sample_kmeans(
  input = sampling_input_pctile,
  clusters = round(
    study_areas[study_area_idx, ]$Shape_Area * sampling_zones_ha / 10000
  ),
  use_xy = TRUE,
  sp_pts = TRUE,
  xy_weight = 2,
  candidates = candidates_field,
  min_cluster_size = 50,
  seed = 5082
)

plot(
  as.factor(myclusters$clusters),
  col = carto_pal(9, "Bold"),
  main = "Primary clusters",
  extent = study_areas[study_area_idx, ],
  buffer = TRUE
)
plot(myclusters$points, pch = 21, bg = "white", add = TRUE)
plot(study_areas[study_area_idx, ], add = TRUE)
```

## 6. Add Extra Sampling Points

The alternative sampling points are generated by using k-means clustering on each of the primary clusters. The additional points are selected from the candidate pixels within the cluster, and the number of additional points is proportional to the cluster size. Unlike the primary sampling points, the alternative sampling points are selected only based on the spatial coordinates of the pixels. This ensures an even geographic distribution of sampling locations within the clusters.

For each cluster, the alternative point located closest to the primary sampling point in each zone is omitted to avoid duplicate sampling locations and to keep the number of sampling locations proportional to the cluster size (each zone already contains one primary sampling location).

```{r extra-points}
cluster_ids <- values(myclusters$clusters, mat = FALSE)
cluster_ids <- cluster_ids[!is.na(cluster_ids)]
cluster_areas <- table(cluster_ids)

extra_pts <- list()
list_index <- 1

for (i in seq_len(nrow(myclusters$points))) {
  cid <- myclusters$points$ID[i]
  csize <- as.numeric(cluster_areas[as.character(cid)])

  if (!is.na(csize) && csize >= (100 / sampling_points_ha)) {
    rast_i <- mask(
      sampling_input_pctile,
      mask = myclusters$clusters,
      maskvalue = cid,
      inverse = TRUE
    ) |>
      trim()

    candidates_i <- mask(
      candidates_field,
      mask = myclusters$clusters,
      maskvalue = cid,
      inverse = TRUE
    ) |>
      crop(rast_i)

    distances_i <- mask(
      myclusters$distances,
      mask = myclusters$clusters,
      maskvalue = cid,
      inverse = TRUE
    ) |>
      crop(rast_i)

    clusters_i <- sample_kmeans(
      input = rast_i,
      clusters = round(csize / (100 / 3)),
      use_xy = TRUE,
      only_xy = TRUE,
      xy_weight = 2,
      sp_pts = TRUE,
      candidates = candidates_i,
      weights = sqrt(distances_i)
    )

    extra_pts[[list_index]] <- clusters_i$points |>
      mutate(cluster = cid)

    list_index <- list_index + 1
  }
}

n_extra_sets <- length(extra_pts)
n_extra_sets
```

Larger clusters receive additional points, with higher preference for pixels farther from the primary center.

## 7. Rank and Label All Points

```{r ranking}
if (length(extra_pts) > 0) {
  extra_pts <- do.call(rbind, extra_pts)

  all_pts <- bind_spat_rows(
    myclusters$points,
    extra_pts
  ) |>
    mutate(
      cluster = case_when(
        is.na(cluster) ~ ID,
        .default = cluster
      )
    ) |>
    unique(atts = FALSE)
} else {
  all_pts <- myclusters$points |>
    mutate(cluster = ID)
}

all_pts <- terra::extract(
  myclusters$distances,
  all_pts,
  bind = TRUE,
  ID = FALSE
) |>
  arrange(cluster, lyr1) |>
  group_by(cluster) |>
  mutate(
    rank = rank(lyr1, ties.method = "first"),
    suffix = index_to_letters(rank),
    label = paste0(cluster, suffix)
  ) |>
  ungroup()

all_pts |>
  as.data.frame() |>
  select(cluster, rank, label, lyr1) |>
  head(15)
```

The point closest to each cluster center is rank 1 (for example 7a), then rank 2 (7b), and so on.

## 8. Final Presentation Map

```{r final-map, out.width = "80%"}
plot(
  sampling_input_field[[2]],
  col = gray.colors(100, start = 0, end = 1),
  buffer = TRUE,
  main = "Final ranked sampling points"
)
plot(
  as.polygons(myclusters$clusters),
  1,
  add = TRUE,
  alpha = 0.25,
  col = carto_pal(9, "Bold"),
  legend = FALSE,
  border = "black"
)
plot(study_areas[study_area_idx, ], add = TRUE, lty = "32")
plot(all_pts, pch = 21, bg = "white", add = TRUE)
text(
  all_pts,
  all_pts$label,
  cex = 0.7,
  col = "black",
  pos = 3,
  hc = "white",
  hw = 0.1,
  halo = TRUE
)
```

## 9. Quick Summary Metrics

```{r summary}
summary_tbl <- all_pts |>
  as.data.frame() |>
  count(cluster, name = "n_points") |>
  arrange(cluster)

summary_tbl

cat("Total points:", nrow(all_pts), "\n")
cat("Primary points:", nrow(myclusters$points), "\n")
cat("Additional points:", nrow(all_pts) - nrow(myclusters$points), "\n")
```

## Notes

-   This report is designed for one study area at a time using params\$study_area_idx.
-   Use the same random seed to keep results reproducible.
-   If data paths change across machines, only params\$data_dir needs to be updated.
