---
title: "Ranked Sampling Workflow for UDKIK"
author: "Anders Bjørn Møller, SEGES"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    fig_width: 10
    fig_height: 7
params:
  data_dir: "C:/Users/anbm/OneDrive - SEGES Innovation PS/UDKIK/Data/"
  study_area_idx: 3
---

## Purpose

This document presents the ranked sampling workflow used in the demo script.
The goal is to place one primary sampling point per cluster, add extra points in larger clusters, and rank the points by distance to the cluster center.

## Workflow Overview

The workflow follows these stages:

1. Read study area and raster inputs.
2. Crop and mask rasters to one field.
3. Transform input layers to percentiles.
4. Clamp outliers to reduce extreme values.
5. Build a candidate mask for valid sample locations.
6. Run primary k-means sampling with spatial weighting.
7. Add extra points inside large clusters.
8. Rank points within each cluster by center distance.
9. Plot final labeled points.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(terra)
library(samplekmeans)
library(tidyverse)
library(dplyr)
library(tidyterra)
library(vctrs)
library(rcartocolor)

set.seed(124)

index_to_letters <- function(n) {
  sapply(n, function(k) {
    out <- ""
    while (k > 0) {
      k <- k - 1
      out <- paste0(letters[(k %% 26) + 1], out)
      k <- k %/% 26
    }
    out
  })
}
```

## Parameters and Input Validation

```{r preflight}
dir_data <- params$data_dir
study_area_idx <- params$study_area_idx

study_area_path <- file.path(dir_data, "Study_areas", "Study_areas.shp")
sampling_paths <- file.path(
  dir_data,
  c(
    "Sampling_input/dhm2015_terraen_10m.tif",
    "Sampling_input/peat_probability_2025_resample.tif",
    "Sampling_input/vdtochn.tif"
  )
)

checks <- tibble::tibble(
  item = c("Study areas", "Terrain", "Peat probability", "Hydrology"),
  path = c(study_area_path, sampling_paths),
  exists = file.exists(c(study_area_path, sampling_paths))
)

checks

can_run <- all(checks$exists)

if (!can_run) {
  cat("Missing one or more required data files. Update params$data_dir and knit again.\n")
}
```

```{r stop-if-missing, include=FALSE}
if (!can_run) {
  knitr::knit_exit()
}
```

## 1. Study Area Selection

```{r study-area, out.width = "60%"}
study_areas <- vect(study_area_path)

values(study_areas)

plot(study_areas[study_area_idx, ], main = paste("Study area", study_area_idx))
```

This selects one study area polygon for demonstration.

## 2. Load and Subset Raster Inputs

```{r load-rasters, out.width = "80%"}
sampling_input <- rast(sampling_paths)

sampling_input_field <- crop(
  sampling_input,
  study_areas[study_area_idx, ]
) |>
  mask(
    study_areas[study_area_idx, ],
    touches = FALSE
  )

sampling_input_field <- ifel(
  sum(is.na(sampling_input_field)) == 0,
  sampling_input_field,
  NA
)

plot(sampling_input_field)
```

Cropping and masking keep only pixels inside the selected field.

## 3. Percentile Transformation and Outlier Clamp

```{r pctile-and-clamp, out.width = "80%"}
sampling_input_pctile <- sapp(
  sampling_input_field,
  function(x) {
    x_ecdf <- ecdf(values(x))
    app(x, x_ecdf)
  }
)

field_means <- global(sampling_input_pctile, "mean", na.rm = TRUE) |>
  unlist()
field_sds <- global(sampling_input_pctile, "sd", na.rm = TRUE) |>
  unlist()

sampling_input_pctile <- sampling_input_pctile |>
  clamp(
    lower = field_means - field_sds * 3,
    upper = field_means + field_sds * 3
  )

plot(sampling_input_pctile)
```

Percentiles normalize layers to a comparable scale. Clamping reduces the effect of extreme values.

## 4. Candidate Sampling Pixels

```{r candidates, out.width = "60%"}
candidates_field <- ifel(
  is.na(sum(sampling_input_field)),
  NA,
  1
) |>
  focal(w = 3, "mean") |>
  mask(
    study_areas[study_area_idx, ] |> as.lines(),
    inverse = TRUE
  )

plot(candidates_field, main = "Candidate sampling pixels")
plot(study_areas[study_area_idx, ], add = TRUE)
```

The candidate mask avoids border effects and excludes unsuitable edge cells.

## 5. Primary Cluster Sampling

```{r primary-clustering, out.width = "60%"}
myclusters <- sample_kmeans(
  input = sampling_input_pctile,
  clusters = round(study_areas[study_area_idx, ]$Shape_Area / 10000),
  use_xy = TRUE,
  sp_pts = TRUE,
  xy_weight = 2,
  candidates = candidates_field,
  min_cluster_size = 50
)

plot(
  as.factor(myclusters$clusters),
  col = carto_pal(9, "Bold"),
  main = "Primary clusters"
)
plot(myclusters$points, pch = 21, bg = "white", add = TRUE)
plot(study_areas[study_area_idx, ], add = TRUE)
```

One primary point is selected per cluster to ensure broad within-field coverage.

## 6. Add Extra Points in Large Clusters

```{r extra-points}
cluster_ids <- values(myclusters$clusters, mat = FALSE)
cluster_ids <- cluster_ids[!is.na(cluster_ids)]
cluster_areas <- table(cluster_ids)

extra_pts <- list()
list_index <- 1

for (i in seq_len(nrow(myclusters$points))) {
  cid <- myclusters$points$ID[i]
  csize <- as.numeric(cluster_areas[as.character(cid)])

  if (!is.na(csize) && csize > 50) {
    rast_i <- mask(
      sampling_input_pctile,
      mask = myclusters$clusters,
      maskvalue = cid,
      inverse = TRUE
    ) |>
      trim()

    candidates_i <- mask(
      candidates_field,
      mask = myclusters$clusters,
      maskvalue = cid,
      inverse = TRUE
    ) |>
      crop(rast_i)

    distances_i <- mask(
      myclusters$distances,
      mask = myclusters$clusters,
      maskvalue = cid,
      inverse = TRUE
    ) |>
      crop(rast_i)

    clusters_i <- sample_kmeans(
      input = rast_i,
      clusters = round(csize / (100 / 3)),
      use_xy = TRUE,
      only_xy = TRUE,
      xy_weight = 2,
      sp_pts = TRUE,
      candidates = candidates_i,
      weights = sqrt(distances_i)
    )

    extra_pts[[list_index]] <- clusters_i$points |>
      mutate(cluster = cid)

    list_index <- list_index + 1
  }
}

n_extra_sets <- length(extra_pts)
n_extra_sets
```

Larger clusters receive additional points, with higher preference for pixels farther from the primary center.

## 7. Rank and Label All Points

```{r ranking}
if (length(extra_pts) > 0) {
  extra_pts <- do.call(rbind, extra_pts)

  all_pts <- bind_spat_rows(
    myclusters$points,
    extra_pts
  ) |>
    mutate(
      cluster = case_when(
        is.na(cluster) ~ ID,
        .default = cluster
      )
    ) |>
    unique(atts = FALSE)
} else {
  all_pts <- myclusters$points |>
    mutate(cluster = ID)
}

all_pts <- terra::extract(
  myclusters$distances,
  all_pts,
  bind = TRUE,
  ID = FALSE
) |>
  arrange(cluster, lyr1) |>
  group_by(cluster) |>
  mutate(
    rank = rank(lyr1, ties.method = "first"),
    suffix = index_to_letters(rank),
    label = paste0(cluster, suffix)
  ) |>
  ungroup()

all_pts |>
  as.data.frame() |>
  select(cluster, rank, label, lyr1) |>
  head(15)
```

The point closest to each cluster center is rank 1 (for example 7a), then rank 2 (7b), and so on.

## 8. Final Presentation Map

```{r final-map, out.width = "80%"}
plot(
  sampling_input_field[[2]],
  col = gray.colors(100, start = 0, end = 1),
  buffer = TRUE,
  main = "Final ranked sampling points"
)
plot(
  as.polygons(myclusters$clusters),
  1,
  add = TRUE,
  alpha = 0.25,
  col = carto_pal(9, "Bold"),
  legend = FALSE,
  border = "black"
)
plot(study_areas[study_area_idx, ], add = TRUE, lty = "32")
plot(all_pts, pch = 21, bg = "white", add = TRUE)
text(
  all_pts,
  all_pts$label,
  cex = 0.7,
  col = "black",
  pos = 3,
  hc = "white",
  hw = 0.1,
  halo = TRUE
)
```

## 9. Quick Summary Metrics

```{r summary}
summary_tbl <- all_pts |>
  as.data.frame() |>
  count(cluster, name = "n_points") |>
  arrange(cluster)

summary_tbl

cat("Total points:", nrow(all_pts), "\n")
cat("Primary points:", nrow(myclusters$points), "\n")
cat("Additional points:", nrow(all_pts) - nrow(myclusters$points), "\n")
```

## Notes

- This report is designed for one study area at a time using params$study_area_idx.
- Use the same random seed to keep results reproducible.
- If data paths change across machines, only params$data_dir needs to be updated.
